### 2024-11-08 docs/paper.tex
Fixed Jo's comments and revised first paragraph

\section{Introduction}
Housing is the largest asset in U.S. households' portfolios. While homeownership is seen as central to the ``American Dream'' \citep{goodman2018homeownership}, households face different barriers to ownership. In particular, around 30\% of American first-time homebuyers received downpayment assistance from their parents between 2009 and 2016. Parental downpayment assistance averaged \$48,000 among receiving households.\footnote{Data from the Survey of Household Economic Decisionmaking and the Transfer Supplement of the 2013  Panel Study of Income Dynamics (see Section \ref{sec:data}).}  It is important to understand how transfers shape households' ability to access and maintain ownership, given  housing’s importance to household portfolios, economic outcomes \citep{sodini2023identifying}, and business cycles \citep{Kaplan2020,favilukis2017macroeconomic}.

This paper quantifies the importance of parental transfers for young adults' homeownership and investigates how policies lowering homeownership barriers interact with parental transfers. Previous empirical studies highlight that parental transfer receipt is associated with an increase in the probability of becoming a homeowner, mainly by relaxing credit constraints.\citep[see e.g.,][]{Lee2018,Blickle2019,wold2024housing}. Additionally. parental transfers help households maintain homeownership, a mechanism that has received less attention \citep[see e.g.,][]{bond2021role}. However, while these studies show that receiving transfers causally increases homeownership, they do not address the overall importance of parental transfers. For example, households may anticipate future transfers and subsequently save less. In this case, even if transfers increase the likelihood of homeownership all else equal, they may not affect aggregate homeownership.

@@

To quantify the impact of parental transfers on homeownership, I then combine two distinct models: housing models, which ignore parental transfers, and models of parental transfers, which ignore housing. I build a life-cycle overlapping generations model with altruistic parents and a rent-or-own decision. Parents and children interact without commitment through both inter-vivos transfers and end-of-life bequests. In models of altruism without housing (e.g., \citealp{Altonji1997a,Barczyk2014}), the transfer motive is to increase the child's consumption when they are borrowing constrained, thus having high marginal utility of wealth. The inclusion of a frictional housing market generates two new transfer motives. First, transfers can directly alleviate credit constraints, enabling wealthier parents to help their children meet downpayment requirements. Second, future transfers provide partial insurance against future income shocks, reducing risks associated with the large illiquid investment of homeownership. Moreover, the illiquidity of housing serves as a commitment device enabling children to secure future transfers. As in all transfer models, the expectation of future transfers reduces children’s savings motives, potentially lowering the homeownership rate.

@@

I estimate the model by matching data on homeownership, wealth, and transfers from the Panel Survey of Income Dynamics (PSID). To quantify the importance of transfers, I find the counterfactual homeownership rate without parental transfers. Without parental transfers, the model simplifies to a standard life-cycle model with housing. I find that parental transfers account for 14 percentage points (29\%) of the young homeownership rate.

Next, I evaluate policies to increase homeownership and their effect on the link between parental wealth and children's housing outcomes. The results show that many policies mainly benefit households with wealthy parents. Relaxing mortgage constraints, for example by lowering minimum downpayments, strengthens the link between parental wealth and children's housing outcomes since households with wealthier parents find homeownership more attractive, all else equal. Other policies that lower barriers to entry, such as reducing prices or purchase costs, increase homeownership for all households, though mainly among those with wealthier parents. However, increasing liquidity by reducing sales costs weakens the link between parental wealth and children’s housing outcomes, as the partial insurance provided by parental wealth becomes less important.

Next, I study how housing illiquidity affects the strategic behavior arising in the absence of commitment. I find that 27\% of young households prefer to face sales costs on housing. These households, who have wealthy parents, use sales costs as a commitment device to maintain low liquid wealth in the next period, effectively becoming ``wealthy hand-to-mouth''. Lower liquid wealth implies higher marginal propensities to consume (MPC), and hence more parental transfers. As an illustrative example, imagine that a child buys a house and brings no liquid wealth to the next period. If the parent transfer in the next period, the child will either face very low consumption (with a correspondingly high marginal value of wealth) or be forced to liquidate the house and incur sales costs. The parent dislikes both outcomes, and may find it optimal to transfer enough to keep the child in the home. This driver of preference for illiquidity differs from those studied in the literature, such as temptation or time inconsistency \citep{attanasio2024temptation,laibson1997golden}. Furthermore, this preference for illiquidity provides a novel explanation for the existence of high-MPC households \cite[see e.g.,][]{kaplan2022marginal}.

Finally, I quantify the contribution of parental transfers to the Black-White homeownership gap: in the U.S., young White households are twice as likely to be homeowners as Black households. I adapt the methodology of \cite{Ashman2020} and \cite{aliprantis2022dynamics}, who study the racial wealth gap, and recalibrate the income process to capture racial differences. I find that parental transfers account for 7 percentage points (14\%) of the Black-White homeownership gap.

@@

By introducing parental transfers to models of housing, my paper contributes to the literature studying the determinants of homeownership over the life cycle. Recent papers focus on marriage and family formation \citep{Fisher2011,Chang2024,Khorunzhina2019}, housing demand in old age \citep{McGee2019,Barczyk2020a}, and changing borrowing constraints \citep{Paz-Pardo2019,Mabille2020}. These studies highlight the importance of credit constraints and minimum down payments in constraining demand for owner-occupied housing. However, by considering parent-child interactions, I show a dichotomy based on parent wealth in the importance of constraints. For households with poorer parents, the sales and purchase costs decrease ownership by making it riskier to own. This risk is less important for households with richer parents. making mortgage credit constraints more important. My results highlight  that policies increasing ownership through subsidies to first-time buyers will lead to increased housing inequality. My model-based results also complement some of the empirical research on the effect of parental resources and housing outcomes \citep[see e.g.,][ for recent work]{wold2024housing,daysal2023intergenerational,benetton2022dynastic} by quantifying the contribution of transfers to homeownership.

I contribute to the literature on altruistic households who interact without commitment by studying how illiquid housing affects the commitment problem. In these models children undersave to increase parental transfers (e.g., \citealp{Altonji1997a,Boar2018,Barczyk2014,Chu2020}). However, adjustment costs on housing impose future expenditure commitments \citep{Chetty2007,Shore2010}. The illiquidity of housing allows households to commit to having low liquid wealth, which increases future transfers. The paper most closely related to mine is \cite{Barczyk2020a}, which studies the mirror problem: how the homeownership of retired parents affects the economic behavior of the elderly and their adult children. In their model, the parent’s home serves as a commitment device, encouraging wealth bequests from the parent and informal care from the adult offspring.

Finally, I contribute to the literature on parental transfers and life-cycle outcomes by focusing on how parental transfers to young adult households influence their wealth accumulation through homeownership. My results show that households with wealthier parents are willing to buy sooner, take on higher leverage, and hold less liquid precautionary savings. The previous literature has generally focused on parental investment in children's human capital (e.g., \citealp{Lee2019,Daruich2018,gilraine2023public}) or transfers from adult children to retired parents (e.g., \citealp{Mommaerts2016, Barczyk2018, Barczyk2020a}), instead of the effect of parental wealth on household portfolio choices.

>>>>

\section{Introduction}
Housing is the largest asset in U.S. households portfolios. While homeownership is seen as central to the ``American Dream'' \citep{goodman2018homeownership}, households face different barriers to ownership. Between 2009 and 2016, around 30\% of American first-time homebuyers received parental assistance for their down payment, with such assistance averaging \$48,000 per receiving household.\footnote{All data from the Survey of Household Economic Decisionmaking and the Panel Study of Income Dynamics (see Section \ref{sec:data}).} This assistance highlights disparities in access, as only those with sufficiently wealthy parents can expect substantial transfers. Understanding how these transfers shape households' ability to access and maintain ownership, given  housing’s importance to household portfolios, economic outcomes \citep{sodini2023identifying}, and business cycles \citep{Kaplan2020,favilukis2017macroeconomic}.

This paper quantifies the importance of parental transfers for young adults' homeownership and investigates how policies lowering homeownership barriers interact with parental transfers. Previous empirical studies highlight that parental transfer receipt is associated with an increase in the probability of becoming a homeowner, mainly by relaxing credit constraints \citep[see e.g.,][]{Lee2018,Blickle2019,wold2024housing}. Additionally, parental transfers help households maintain homeownership, a mechanism that has received less attention \citep[see e.g.,][]{bond2021role}. However, while these studies show that receiving transfers causally increases homeownership, they do not address the overall importance of parental transfers. For example, households may anticipate future transfers and subsequently save less. In this case, even if transfers increase the likelihood of homeownership all else equal, they may not affect aggregate homeownership.

@@

To quantify the impact of parental transfers on homeownership, I then combine two distinct models: housing models, which ignore parental transfers, and models of parental transfers, which ignore housing. I build a life-cycle overlapping generations model with altruistic parents and a rent-or-own decision. Parents and children interact without commitment through both inter-vivos transfers and end-of-life bequests. In models of altruism without housing (e.g., \citealp{Altonji1997a,Barczyk2014}), the transfer motive is to increase the child's consumption when they are borrowing constrained, thus having high marginal utility of wealth. The inclusion of a frictional housing market generates two new transfer motives. First, transfers can directly alleviate credit constraints, enabling wealthier parents to help their children meet downpayment requirements. Second, future transfers provide partial insurance against future income shocks, reducing risks associated with the large illiquid investment of homeownership. Moreover, the illiquidity of housing serves as a commitment device enabling children to secure future transfers. As in all transfer models, the expectation of future transfers reduces children’s saving motives, potentially lowering the homeownership rate.

@@

I estimate the model by matching data on homeownership, wealth, and transfers from the Panel Study of Income Dynamics (PSID). To quantify the importance of transfers, I find the counterfactual homeownership rate without parental transfers. Without parental transfers, the model simplifies to a standard life-cycle model with housing. I find that parental transfers account for 14 percentage points (29\%) of the homeownership rate among young households.

I then use the model to evaluate policies aiming to increase homeownership. I find that such policies disproportionately benefit households with wealthy parents. Relaxing mortgage constraints, for example by lowering minimum downpayments, strengthens the link between parental wealth and children's housing outcomes since more households with wealthy parents are on the margin to become homeowners. Non-mortgage policies that lower barriers to entry, such as reducing prices or purchase costs, increase homeownership for all households, though disproportionately among those with wealthier parents. However, increasing liquidity by reducing sales costs weakens the link between parental wealth and children’s housing outcomes, as the partial insurance provided by parental wealth becomes less important.

Next, I study how housing illiquidity affects the strategic behavior arising in the absence of commitment. I find that 27\% of young households prefer that housing is illiquid---due to sales costs---rather than liquid. These households have wealthy parents and use the sales costs as a commitment device to maintain low liquid wealth in the next period, effectively becoming ``wealthy hand-to-mouth''. Lower liquid wealth implies higher marginal propensities to consume (MPC), and hence more parental transfers. As an illustrative example, consider a child who buys a house and brings no liquid wealth to the next period. If the parent does not transfer in the next period, the child will either face very low consumption (with a correspondingly high marginal value of wealth) or be forced to liquidate the house and incur sales costs. The parent dislikes both outcomes, and may find it optimal to transfer enough to keep the child in the home. This driver of preference for illiquidity differs from those studied in the literature, such as temptation or time inconsistency \citep{attanasio2024temptation,laibson1997golden}. Furthermore, that parental wealth generates preference for illiquidity provides a novel explanation for the existence of high-MPC households \cite[see e.g.,][]{kaplan2022marginal}.

Finally, I quantify the contribution of parental transfers to the Black-White homeownership gap of 50\% among young adults. I adapt the methodology of \cite{Ashman2020} and \cite{aliprantis2022dynamics}, who study the racial wealth gap, and recalibrate the income process to capture racial differences. I find that parental transfers account for 7 percentage points (14\%) of the Black-White homeownership gap. For context, the Black-White homeownership gap has remained roughly constant since the passing of the Fair Housing Act in 1968.

@@

My paper first contributes to the literature studying the determinants of homeownership over the life cycle. Recent papers focus on marriage and family formation \citep{Fisher2011,Chang2024,Khorunzhina2019}, housing demand in old age \citep{McGee2019,Barczyk2020a}, and changing borrowing constraints \citep{Paz-Pardo2019,Mabille2020}. These studies highlight the importance of credit constraints and minimum down payments in constraining demand for owner-occupied housing. By considering parent-child interactions, I show how parental wealth affects the relative importance of constraints. For households with poorer parents, the sales and purchase costs decrease ownership by making it riskier to own. This risk is smaller  for households with richer parents, making mortgage credit constraints relatively more important. My results highlight that policies increasing ownership through subsidies to first-time buyers---a common approach in many cities---will lead to increased housing inequality. My model-based results also complement some of the empirical research on the effect of parental resources and housing outcomes \citep[see e.g.,][ for recent work]{wold2024housing,daysal2023intergenerational,benetton2022dynastic} by quantifying the contribution of transfers to homeownership.

My paper also contributes to the literature on altruistic households who interact without commitment by studying how illiquid housing affects the commitment problem. In these models children have an incentive to undersave to increase parental transfers (e.g., \citealp{Altonji1997a,Boar2018,Barczyk2014,Chu2020}). However, the illiquidity of housing imposes future expenditure commitments \citep{Chetty2007,Shore2010}. The illiquidity of housing allows households to commit to having low liquid wealth, which increases future transfers. The paper most closely related to mine is \cite{Barczyk2020a}, which studies the mirror problem: how the homeownership of retired parents affects the economic behavior of the elderly and their adult children. In their model, the parent’s home serves as a commitment device, encouraging wealth bequests from the parent and informal care from the adult offspring.

Finally, the paper contributes to the literature on parental transfers and life-cycle outcomes by focusing on how parental transfers to young adult households influence their wealth accumulation through homeownership. My results show that households with wealthier parents are willing to buy sooner, take on higher leverage, and hold less liquid precautionary savings. The previous literature has generally focused on parental investment in children's human capital (e.g., \citealp{Lee2019,Daruich2018,gilraine2023public}) or transfers from adult children to retired parents (e.g., \citealp{Mommaerts2016, Barczyk2018, Barczyk2020a}), instead of the effect of parental wealth on household portfolio choices.
